\bf A Comparison of Parametric and Permutation Tests for Regression Analysis of Randomized Control Trials

### 2017-01-31 ms/SBR-template.tex
Reformat tables nicely

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL 
\usepackage{color}
\newcommand{\todo}[1]{{\color{red}{TO DO: \sc #1}}}

@@

Table~\ref{tab:normal_power} displays the size of the test and the power at level $0.05$.
The three tests based on linear models have slightly higher than nominal level, though the margin is within two standard errors of $0.05$.
(The number of tests rejected under the null in 1000 trials has a binomial distribution.
If the true level is $0.05$, then the standard error is $\sqrt{0.05 \times 0.95/1000}$.)
These numbers show that actually, for $\alpha=0.05$, the stratified permutation test does not lose a large amount of power when the effect is isolated in a single stratum.
In this case, the permutation test may have slightly higher power than the parametric ANCOVA at the small significance levels used in practice.
\begin{figure}[h]
\centering
\includegraphics[width = \textwidth]{fig/normal_simulation_power.pdf}
\caption{Empirical power curves for the Gaussian simulated data}
\label{fig:normal_sim_power}
\end{figure}
\begin{center}
\input{fig/normal_simulation_power_summary.tex}
\end{center}

Our second set of simulations used the same design as the previous ones, but modified the sizes of the treatment groups.
Within each stratum, 4 individuals received one treatment and 12 received the other.
This mimics real-world experiments, where the more expensive treatment is administered less frequently than the other.
It is well-known that both parametric and nonparametric tests have higher than nominal level when there is heterogeneous variance, and this effect is exacerbated when group sizes are unequal (\cite{glass_consequences_1972}, \cite{zimmerman_two_2006}).
Here, the data have the same variance so we would expect power, not the test level, to be an issue.
(See the supplementary Gaussian simulations for a discussion of heterogeneous variances.)
We omit a plot of the power curves from these simulations, as the patterns appear similar to those in Figure~\ref{fig:normal_sim_power}.
Table~\ref{tab:imbalanced_power} displays the value of these power curves and the size of the test at level $0.05$.
In this case, the permutation tests (except for the simple stratified permutation) have slightly higher power than the ANCOVA, but the difference is not substantial.
%\begin{figure}
%\centering
%\includegraphics[width = \textwidth]{fig/imbalanced_simulation_power}
%\caption{Empirical power curves for the Gaussian simulated data with imbalanced group sizes}
%\label{fig:imbalanced_sim_power}
%\end{figure}
\begin{center}
\input{fig/imbalanced_simulation_power_summary.tex}
\end{center}

@@

\begin{center}
\input{fig/skewed_simulation_power_summary.tex}
\end{center}

It makes sense that in all of these simulations, the permutation test using the change scores as the response measure is more powerful than the permutation test using the raw outcomes.
The baseline and outcome are highly correlated, so the change score has lower variance the outcome alone.
In additional simulations, 
we modified the data-generating process given by Equation~\ref{eqn:dgp} so that the outcome and baseline had correlation of $0.25$ (see the supplementary Gaussian simulations for a more detailed treatment).
In this case, the result was reversed: 
the differenced test had low power while the test using the raw outcomes had a power curve closer to the linear model-based tests.
This behavior has been studied before by \cite{frison_repeated_1992}, who recommended to use the difference in outcome and baseline if their correlation is greater than $0.5$ and to use the outcome only if their correlation is less than $0.5$.

@@

\begin{center}
\input{fig/results_pvalue_summary.tex}
\end{center}

>>>>

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL 
\usepackage{color}
 \usepackage{multirow}
\newcommand{\todo}[1]{{\color{red}{TO DO: \sc #1}}}

@@

Table~\ref{tab:normal_power} displays the size of the test and the power at level $0.05$.
The three tests based on linear models have slightly higher than nominal level, though the margin is within two standard errors of $0.05$.
(The number of tests rejected under the null in 1000 trials has a binomial distribution.
If the true level is $0.05$, then the standard error is $\sqrt{0.05 \times 0.95/1000}$.)
These numbers show that actually, for $\alpha=0.05$, the stratified permutation test does not lose a large amount of power when the effect is isolated in a single stratum.
In this case, the permutation test may have slightly higher power than the parametric ANCOVA at the small significance levels used in practice.
\begin{figure}[h]
\centering
\includegraphics[width = \textwidth]{fig/normal_simulation_power.pdf}
\caption{Empirical power curves for the Gaussian simulated data}
\label{fig:normal_sim_power}
\end{figure}
%\begin{center}
%\input{fig/normal_simulation_power_summary.tex}
%\end{center}

\begin{table}[]
\centering
\label{tab:normal_power}
\begin{tabular}{ll|ccccc}
\multicolumn{2}{c|}{Design}                                                                  & \multicolumn{5}{c}{$p$-value}                                                                                                                                                                                                                                                                                                                                           \\ \hline
Errors                        & \begin{tabular}[c]{@{}l@{}}Treatment \\ Effect\end{tabular} & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Differenced\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
\multirow{3}{*}{Gaussian}     & None                                                        & 0.058                      & 0.042                                                                                & 0.038                                                                                 & 0.062                                                                        & 0.057                                                                        \\
                              & Constant                                                    & 0.886                      & 0.545                                                                                 & 0.858                                                                                 & 0.891                                                                        & 0.892                                                                        \\
                              & Single stratum                                              & 0.159                      & 0.089                                                                                & 0.160                                                                                 & 0.162                                                                        & 0.170                                                                        \\ \hline
\multirow{3}{*}{Heavy-tailed} & None                                                        & 0.032                      & 0.040                                                                                & 0.045                                                                                 & 0.042                                                                        & 0.040                                                                        \\
                              & Constant                                                    & 0.355                      & 0.240                                                                                & 0.378                                                                                 & 0.400                                                                        & 0.393                                                                        \\
                              & Single stratum                                              & 0.066                      & 0.080                                                                                & 0.082                                                                                 & 0.076                                                                        & 0.080                                                                        \\ \hline
\end{tabular}
\caption{Empirical power at level $0.05$ for Gaussian simulated data.}
\end{table}

Our second set of simulations used the same design as the previous ones, but modified the sizes of the treatment groups.
Within each stratum, 4 individuals received one treatment and 12 received the other.
This mimics real-world experiments, where the more expensive treatment is administered less frequently than the other.
It is well-known that both parametric and nonparametric tests have higher than nominal level when there is heterogeneous variance, and this effect is exacerbated when group sizes are unequal (\cite{glass_consequences_1972}, \cite{zimmerman_two_2006}).
Here, the data have the same variance so we would expect power, not the test level, to be an issue.
(See the supplementary Gaussian simulations for a discussion of heterogeneous variances.)
We omit a plot of the power curves from these simulations, as the patterns appear similar to those in Figure~\ref{fig:normal_sim_power}.
Table~\ref{tab:imbalanced_power} displays the value of these power curves and the size of the test at level $0.05$.
In this case, the permutation tests (except for the simple stratified permutation) have slightly higher power than the ANCOVA, but the difference is not substantial.
%\begin{figure}
%\centering
%\includegraphics[width = \textwidth]{fig/imbalanced_simulation_power}
%\caption{Empirical power curves for the Gaussian simulated data with imbalanced group sizes}
%\label{fig:imbalanced_sim_power}
%\end{figure}
%\begin{center}
%\input{fig/imbalanced_simulation_power_summary.tex}
%\end{center}

\begin{table}[ht]
\centering
\label{tab:imbalanced_power}
\begin{tabular}{ll|ccccc}
\multicolumn{2}{c|}{Design}                                                                 & \multicolumn{5}{c}{$p$-value}                                                                                                                                                                                                                                                                                                                                           \\ \hline
Errors                        & \begin{tabular}[c]{@{}l@{}}Treatment \\ Effect\end{tabular} & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Differenced\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
\multirow{3}{*}{Gaussian}     & None                                                        & 0.046                      & 0.031                                                                                & 0.028                                                                                 & 0.042                                                                        & 0.045                                                                        \\
                              & Constant                                                    & 0.840                      & 0.490                                                                                & 0.810                                                                                 & 0.870                                                                        & 0.860                                                                        \\
                              & Single stratum                                              & 0.120                      & 0.050                                                                                & 0.090                                                                                 & 0.100                                                                        & 0.120                                                                        \\ \hline
\multirow{3}{*}{Heavy-tailed} & None                                                        & 0.034                      & 0.054                                                                                & 0.055                                                                                 & 0.047                                                                        & 0.042                                                                        \\
                              & Constant                                                    & 0.230                      & 0.200                                                                                & 0.270                                                                                 & 0.260                                                                        & 0.270                                                                        \\
                              & Single stratum                                              & 0.070                      & 0.080                                                                                & 0.090                                                                                 & 0.090                                                                        & 0.080                                                                        \\ \hline
\end{tabular}
\caption{Empirical power at level $0.05$ for Gaussian simulated data with imbalanced treatment groups} 
\end{table}

@@

%\begin{center}
%\input{fig/skewed_simulation_power_summary.tex}
%\end{center}
\begin{table}[]
\centering
\label{tab:skewed_power}
\begin{tabular}{ll|ccccc}
\multicolumn{2}{c|}{Design}                                                               & \multicolumn{5}{c}{$p$-value}                                                                                                                                                                                                                                                                                                                                           \\ \hline
Group sizes                 & \begin{tabular}[c]{@{}l@{}}Treatment \\ Effect\end{tabular} & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Differenced\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
\multirow{3}{*}{Balanced}   & None                                                        & 0.036                      & 0.025                                                                                & 0.021                                                                                 & 0.035                                                                        & 0.037                                                                        \\
                            & Constant                                                    & 0.974                      & 0.143                                                                                & 0.853                                                                                 & 0.972                                                                        & 0.977                                                                        \\
                            & Single stratum                                              & 0.422                      & 0.050                                                                                & 0.275                                                                                 & 0.429                                                                        & 0.424                                                                        \\ \hline
\multirow{3}{*}{Imbalanced} & None                                                        & 0.038                      & 0.024                                                                                & 0.019                                                                                 & 0.037                                                                        & 0.039                                                                        \\
                            & Constant                                                    & 0.971                      & 0.147                                                                                & 0.860                                                                                 & 0.970                                                                        & 0.972                                                                        \\
                            & Single stratum                                              & 0.190                      & 0.070                                                                                & 0.100                                                                                 & 0.190                                                                        & 0.190                                                                        \\ \hline
\end{tabular}
\caption{Empirical power at level $0.05$ for discrete, skewed simulated data} 
\end{table}
It makes sense that in all of these simulations, the permutation test using the change scores as the response measure is more powerful than the permutation test using the raw outcomes.
The baseline and outcome are highly correlated, so the change score has lower variance the outcome alone.
In additional simulations, 
we modified the data-generating process given by Equation~\ref{eqn:dgp} so that the outcome and baseline had correlation of $0.25$ (see the supplementary Gaussian simulations for a more detailed treatment).
In this case, the result was reversed: 
the differenced test had low power while the test using the raw outcomes had a power curve closer to the linear model-based tests.
This behavior has been studied before by \cite{frison_repeated_1992}, who recommended to use the difference in outcome and baseline if their correlation is greater than $0.5$ and to use the outcome only if their correlation is less than $0.5$.

@@

%
%\begin{center}
%\input{fig/results_pvalue_summary.tex}
%\end{center}

\begin{table}[]
\centering
\label{tab:clinical_pvalues}
\begin{tabular}{l|cccc}
Endpoint      & \multicolumn{1}{l}{ANCOVA} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Stratified\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}LM\\ Permutation\end{tabular}} & \multicolumn{1}{l}{\begin{tabular}[c]{@{}l@{}}Freedman-\\ Lane\end{tabular}} \\ \hline
heart\_freq   & 0.035                      & 0.006                                                                                & 0.080                                                                        & 0.082                                                                        \\
regurg\_freq  & 0.136                      & 0.118                                                                                & 0.280                                                                        & 0.220                                                                        \\
dysp\_freq    & 0.565                      & 0.948                                                                                & 0.616                                                                        & 0.592                                                                        \\
daily\_heart & 0.032                      & 0.004                                                                                & 0.056                                                                        & 0.068                                                                        \\
daily\_regur  & 0.142                      & 0.174                                                                                & 0.286                                                                        & 0.246                                                                        \\
daily\_hrdq   & 0.043                      & 0.012                                                                                & 0.088                                                                        & 0.098                                                                        \\
daily\_dysp   & 0.582                      & 0.810                                                                                & 0.756                                                                        & 0.722                                                                        \\
\hline
\end{tabular}
\caption{Comparison of p-values from four tests, for each continuous endpoint.} 
\end{table}

### 2017-01-31 ms/fig/normal_simulation_power_summary.tex
Reformat tables nicely

\begin{table}[ht]
\centering
\begin{tabular}{p{1.15in}|p{0.7in}|p{0.6in}p{0.8in}p{0.8in}p{0.8in}p{0.75in}}
  \hline
Treatment & Errors & ANOVA & Stratified Permutation & Differenced Permutation & LM Permutation & Freedman-Lane \\ 
  \hline
Constant Additive Effect & Gaussian & 0.89 & 0.55 & 0.86 & 0.89 & 0.89 \\ 
  Single Stratum Effect & Gaussian & 0.16 & 0.09 & 0.16 & 0.16 & 0.17 \\ 
  Constant Additive Effect & Heavy-tailed & 0.35 & 0.24 & 0.38 & 0.40 & 0.39 \\ 
  Single Stratum Effect & Heavy-tailed & 0.07 & 0.08 & 0.08 & 0.08 & 0.08 \\ 
   \hline
\end{tabular}
\caption{Empirical power at level $0.05$ for Gaussian simulated data} 
\label{tab:normal_power}
\end{table}

>>>>

\begin{table}[ht]
\centering
\begin{tabular}{p{1.15in}|p{0.7in}|p{0.6in}p{0.8in}p{0.8in}p{0.8in}p{0.75in}}
  \hline
Errors & Treatment & ANCOVA & Stratified Permutation & Differenced Permutation & LM Permutation & Freedman-Lane \\ 
  \hline
Gaussian & No Effect & 0.058 & 0.042 & 0.038 & 0.062 & 0.057 \\ 
  Gaussian & Constant Additive Effect & 0.886 & 0.545 & 0.858 & 0.891 & 0.892 \\ 
  Gaussian & Single Stratum Effect & 0.159 & 0.089 & 0.160 & 0.162 & 0.170 \\ 
   \hline
Heavy-tailed & No Effect & 0.032 & 0.040 & 0.045 & 0.042 & 0.040 \\ 
  Heavy-tailed & Constant Additive Effect & 0.355 & 0.240 & 0.378 & 0.400 & 0.393 \\ 
  Heavy-tailed & Single Stratum Effect & 0.066 & 0.080 & 0.082 & 0.076 & 0.080 \\ 
   \hline
\end{tabular}
\caption{Empirical power at level $0.05$ for Gaussian simulated data} 
\label{tab:normal_power}
\end{table}
